---
title: "Week7 Exercises"
author: "David Post"
date: "5/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## <span style="color:cornflowerblue">In-class Exercise 1</span>   {.smaller .build}

Consider the following matrix
```{r}
library(datasets)
data(state)
state.x77 %>% summary()
```

## <span style="color:cornflowerblue">In-class Exercise 1</span>   {.smaller .build}


1. Write a function that takes a numeric matrix and transforms each column by substracting by the column mean and dividing by the column standard deviation.  Use this function on the `state.x77` matrix

2. Combining `apply` and `lapply`  write a function to obtain 10 subsamples of from `state.x77` by selecting 20 rows at random each time, and  calculate the column means for each of these subsamples.

3. In problem 2, replace `lapply` by `sapply`

### 1

```{r}
apply(state.x77, 1, FUN=function(x) (x - mean(x))/sd(x))
```

### 2

```{r}
lapply(1:10, function(x){
  apply(state.x77[sample(1:nrow(state.x77),20),],MARGIN=2, mean)
})
```

### 3

```{r}
sapply(1:10, function(x){
  apply(state.x77[sample(1:nrow(state.x77),20),],MARGIN=2, mean)
})
```
