---
title: 'Week 3: R Objects and Intro to Data Vis'
author: "Daniel Taylor-Rodriguez"
date: "April 14, 2020"
output:
  ioslides_presentation:
   incremental: true
subtitle: STAT 399
---

<style>
.forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>

```{r, message=FALSE, warning=FALSE, include=FALSE}
options(
  htmltools.dir.version = FALSE, # for blogdown
  width = 80,
  tibble.width = 80
)

knitr::opts_chunk$set(
  fig.align = "center"
)

htmltools::tagList(rmarkdown::html_dependency_font_awesome())

library(dplyr)
library(purrr)
library(tidyr)
library(magrittr)
library(ggplot2)
```



## Plan for this Week {.build}

### <span style="color:cornflowerblue"> Part 1: wrap up base R </span>

1. Finish subsetting (matrices, factors and data.frames)

2. Subsetting and assignment

### <span style="color:cornflowerblue"> Part 2: modern R basics </span>

3. Visualization in base R 

4. Visualization in ggplot2 (intro)


## Peer Review: Grading Rubric 


- Click [here](https://drive.google.com/drive/folders/0B7QLO8F4QBfXNnJQYzlqdmp4VGs?usp=sharing) and look for the folder with your name.  There you will find the two files (`Rmd` and `html`) you are provide a peer review for. <span style="color:orange"> 

- **You must be signed with your `edu.pdx` account to access the document.**</span>


- Type your *peer evaluation* in word making reference to the line in the Rmd fileor to the specific homework problem number to which you are alluding to.



## Peer Review: Grading Rubric {.smaller}

<span style="color:forestgreen">**Again, be constructive and considerate.** </span><span style="color:orange">**Use a continuous scale between 0 and 3, using the following as reference:**</span> 

0 - No homework turned in.


1 - Turned in but low effort, poorly presented with nonfunctional code and ignoring directions.


2 - Decent effort, well presented all code works and followed directions with some minor issues.


3 - Nailed it!


# Data Visualization in R

## Data Visualization in R before `ggplot2` {.build}

- Since R is a software explicitly developed to do statistics R it comes with extensive plotting capabilities by default

- <span style="color:cornflowerblue">**MANY** plotting functions are installed in the `graphics` package, which ships with base R</span>

- Look into the help files for the functions in this package using `library(help = "graphics")`

- Here are a few examples of visualization functions in the `graphics` package:

<div class="centered">
`plot`, `lines`, `points`,`abline`,`boxplot`, `pairs`, `matplot`, `barplot`, `curve`, `dotchart`, `pie`, `rasterImage`, `coplot`, `cdplot`, `mosaicplot`, `polygon`</span>
</div>

## Data visualization before `ggplot2` {.smaller .build}

### `plot`

- `plot` is generic function for plotting R objects

- Functions in R can be designed in such a way that the same function can have a completely different behavior depending on the object it is used with

- Here is a list with some of the objects it can be applied to

```{r}
methods(plot)
```


## Data visualization before `ggplot2` {.smaller .build}

### `plot`

The default form of the function generates *scatterplots*

Its general structure is `plot(x, y, ...)`, where 

- `x` is the variable in the x axis, 

- `y` is the variable in the y axis, and 

- `...` represents other graphical parameters (see `?par` for an extensive list)

Let's do an example to see some of the options

## Data visualization before `ggplot2` {.smaller .build}

### Example: the `cars` dataset

Let's load the built-in data `cars`, which loads as a dataframe, a type of object mentioned earlier. Then, we can look at it in a couple different ways.

`data(cars)` loads this dataframe into the **Global Environment** as a *promise*. Promises are *unevaluated arguments*.

```{r}
str(cars, 5) 
```

## Data visualization before `ggplot2` {.smaller .build}

### Example: the `cars` dataset {.build}

```{r}
head(cars,4) # prints first 4 rows
```

```{r}
summary(cars) # summary stats for each var
```

## Data visualization before `ggplot2` {.smaller}

### `plot` options
```{r, eval=F,echo=T, fig.height=5, fig.width=5}
plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16,
     cex = 1, cex.axis = 1, cex.lab = 1,
     col = "firebrick") 
```


## Data visualization before `ggplot2` {.smaller}

### `plot` options
```{r, eval=T,echo=F, fig.height=4.5, fig.width=4.5}
plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16,
     cex = 1, cex.axis = 1, cex.lab = 1,
     col = "firebrick") 
```




## Data visualization before `ggplot2` {.smaller}

### `plot` options: `type`

```{r, eval=T,echo=F, fig.height=2.5, fig.width=7.5}
par(mfrow=c(1,3))
plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "type = 'p'",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16,
     cex = 1, cex.axis = 1, cex.lab = 1,
     col = "firebrick") 
plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "type = 'l'",
     type = "l",
     lty = 1, lwd = 1,
     pch = 16,
     cex = 1, cex.axis = 1, cex.lab = 1,
     col = "firebrick") 
plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "type = 'b'",
     type = "b",
     lty = 1, lwd = 1,
     pch = 16,
     cex = 1, cex.axis = 1, cex.lab = 1,
     col = "firebrick") 
```

## Data visualization before `ggplot2` {.smaller}

### `plot` options: `cex`
```{r, eval=F,echo=T, fig.height=4.5, fig.width=4.5}
obj.size <- seq(0.5,5,length.out = length(cars$speed))

plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16,
     cex = obj.size, cex.axis = 1, cex.lab = 1,
     col = "firebrick") 
```

## Data visualization before `ggplot2` {.smaller}

### `plot` options: `cex`
```{r, eval=T,echo=F, fig.height=4.5, fig.width=4.5}
obj.size <- seq(0.5,5,length.out = length(cars$speed))

plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16,
     cex = obj.size, cex.axis = 1, cex.lab = 1,
     col = "firebrick") 
```

## Data visualization before `ggplot2` {.smaller}

### `plot` options: `col`
```{r, eval=F,echo=T, fig.height=4.5, fig.width=4.5}
col.vec <- rep(c("firebrick","forestgreen","cornflowerblue"),
               times=c(sum(cars$speed<10), 
                       sum(cars$speed>=10&cars$speed<17),
                       sum(cars$speed>=17)))

plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16,
     cex = 1, cex.axis = 1, cex.lab = 1,
     col = col.vec) 
```


## Data visualization before `ggplot2` {.smaller}

### `plot` options: `col`
```{r, eval=T,echo=F, fig.height=4.5, fig.width=4.5}
col.vec <- rep(c("firebrick","forestgreen","cornflowerblue"),
               times=c(sum(cars$speed<10), 
                       sum(cars$speed>=10&cars$speed<17),
                       sum(cars$speed>=17)))

plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16,
     cex = 1, cex.axis = 1, cex.lab = 1,
     col = col.vec) 
```


## Data visualization before `ggplot2` {.smaller}

### Adding a `legend`
```{r, eval=F,echo=T, fig.height=4.5, fig.width=4.5}
plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16,
     cex = 1, cex.axis = 1, cex.lab = 1,
     col = col.vec) 

legend("topleft",
       bty = "n",
       pch=c(16,16,16),
       col=c("firebrick","forestgreen","cornflowerblue"),
       legend=c("speed<10","10<=speed<17","speed>=17")) 
```


## Data visualization before `ggplot2` {.smaller}

### Adding a `legend`
```{r, eval=T,echo=F, fig.height=4.5, fig.width=4.5}
plot(x = cars$speed, y = cars$dist,
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     type = "p",
     lty = 1, lwd = 1,
     pch = 16,
     cex = 1, cex.axis = 1, cex.lab = 1,
     col = col.vec) 
legend("topleft",
       bty = "n",
       pch=c(16,16,16),
       col=c("firebrick","forestgreen","cornflowerblue"),
       legend=c("speed<10","10<=speed<17","speed>=17"))
```



## In-class exercise 1 {.smaller}

Your task is to play with the `swiss` data set built into R fr 20 mins

- Use `?swiss` to see what things mean in the dataset

- Go to the in-class exercise Rmd document you started working on Tuesday 

- Load the data using `data(swiss)`

- Think of and write down in your Rmd document one or two questions you'd like to explore with these data

- Use the function `plot` to explore your questions and make 2 or 3 nicely formatted plots with with the options we discussed so far (include legends, play with `col`, `cex`, `type`)




## Data visualization before `ggplot2` {.smaller}
### Other plots: `boxplot`
```{r boxplotfn, echo=T, fig.height=3, fig.width=6}
par(mfrow=c(1,2),mai=c(1,0.5,0.1,0.1))
boxplot(decrease ~ treatment, data = OrchardSprays, col = "cornflowerblue",
        log = "y",cex.axis=0.7,cex.lab=0.7,notch=F)
## horizontal=TRUE, switching  y <--> x :
boxplot(decrease ~ treatment, data = OrchardSprays, col = "cornflowerblue",
        log = "x", horizontal=TRUE,cex.axis=0.7,cex.lab=0.7,notch=F)

```


## Data visualization before `ggplot2` {.smaller}
### Other plots: `curve`

```{r curvefn, echo=T, fig.height=2.5, fig.width=7.5}
par(mfrow=c(1,3),mai=c(0.9,0.4,0.1,0.1))
curve(expr=sin, from=-2*pi, to=2*pi, xname = "t",cex.axis=0.7, cex.lab=0.7)
curve(expr=tan, xname = "t", from=-2*pi, to=2*pi, cex.axis=0.7, cex.lab=0.7)
myfn <- function(t){tan(t)*sin(t)}
curve(expr=myfn, xname = "t", from=-2*pi, to=2*pi, cex.axis=0.7, cex.lab=0.7)
```


## Data visualization before `ggplot2` {.smaller}
### Other plots: `hist`
```{r histfn, echo=T, fig.height=3, fig.width=6}
par(mfrow=c(1,2),mai=c(1,0.5,0.1,0.1))
x <- rchisq(1000, df = 4)
hist(x, freq = FALSE, ylim = c(0, 0.2),col="orange",main="")
#hist and curve combined
hist(x, freq = FALSE, ylim = c(0, 0.2),col="orange",main="")
lines(density(x,from=0, to=20), col = "blue3", lty = 1, lwd = 3, add = TRUE)
```



## Data visualization before `ggplot2` {.smaller}
### Other plots: `pairs`

```{r pairsfn, echo=T, fig.align='center',fig.height=4, fig.width=4}
par(mai=c(0.1,0.1,0.1,0.1))
pairs(iris[1:3], cex=0.5, cex.labels = 1,cex.axis=0.7,
      pch = 21, bg = c("red", "green3", "blue")[unclass(iris$Species)])
```


## Data visualization before `ggplot2` {.smaller}
### Other plots: `dotchart`
<div class="columns-2">
```{r}
VADeaths
```

```{r, eval=F, echo=T}
par(mai=c(0.4,0.1,0.4,0.1))
dotchart(VADeaths, bg = "skyblue",
         cex=0.7, cex.axis=0.1,
         main = "Death Rate VA - 1940")
```
<p class="forceBreak"></p>

```{r dotchartfn, eval=T, echo=F, fig.align='center',fig.height=3.5, fig.width=3.5}
par(mai=c(0.4,0.1,0.4,0.1))
dotchart(VADeaths, bg = "skyblue",
         cex=0.7, cex.axis=0.1,
         main = "Death Rate VA - 1940")
```
</div>


## Data visualization before `ggplot2` {.smaller}
### Other plots: `matplot`

```{r matplotfn, echo=T, fig.height=3, fig.width=6}
par(mfrow=c(1,2),mai=c(0.4,0.4,0.1,0.1))
sines <- outer(1:20, 1:4, function(x, y) sin(x / 20 * pi * y))
matplot(sines, pch = 1:4, type = "o", col = rainbow(ncol(sines)),
        cex=0.5, cex.axis=0.7)
matplot(sines, type = "b", pch = 21:23, col = 2:5, bg = 2:5, 
        cex=0.5, cex.axis=0.7, main = "")
```

## Data visualization before `ggplot2` {.smaller}
### Other plots: `barplot`

```{r barplotfn, echo=T, fig.height=3, fig.width=6}
par(mfrow=c(1,2),mai=c(0.9,0.4,0.1,0.1))
barplot(GNP ~ Year, data = longley, cex=0.5, cex.axis=0.7, cex.lab=0.7)
barplot(cbind(Employed, Unemployed) ~ Year, data = longley, cex = 0.5,
        cex.axis=0.7,cex.lab=0.7)
```


## Data visualization before `ggplot2` {.smaller}
### Other plots: `mosaic`

```{r mosaicfn, echo=T, fig.height=4, fig.width=4}
par(mfrow=c(1,1),mai=c(0.9,0.4,1,0.4))
mosaicplot(~ Sex + Age + Survived, data = Titanic, main="",color = TRUE)
```


## Data visualization before `ggplot2` {.smaller}

### <span style="color:cornflowerblue">Figures with multiple panels:</span> `mfrow` (or `mfcol`)
```{r, eval=F,echo=T, fig.height=2.5, fig.width=7.5}
par(mfrow=c(1,3))
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
abline(reg=lm(dist~speed,data=cars),col="forestgreen")
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
lines(lowess(cars),col="cornflowerblue")
```

## Data visualization before `ggplot2` {.smaller}

### <span style="color:cornflowerblue">Figures with multiple panels:</span> `mfrow` (or `mfcol`)
```{r, eval=T,echo=F, fig.height=2.8, fig.width=8.4}
par(mfrow=c(1,3))
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
abline(reg=lm(dist~speed,data=cars),col="forestgreen",lwd=2)
plot(x = cars$speed, y = cars$dist,xlab = "", ylab = "", main = "",
     type = "p", pch = 16, col = "firebrick") 
lines(lowess(cars),col="cornflowerblue",lwd=2)
```

## Data visualization before `ggplot2` {.smaller}

### <span style="color:cornflowerblue">Figures with multiple panels:</span> the `layout` function

```{r, eval=FALSE}
layout(mat, 
       widths = rep.int(1, ncol(mat)),
       heights = rep.int(1, nrow(mat)), 
       respect = FALSE)
```

## Data visualization before `ggplot2` {.smaller}

### <span style="color:cornflowerblue">Figures with multiple panels:</span> the `layout` function

```{r, echo=T,eval=T,fig.height=4,fig.width=4}
nf <- layout(matrix(c(2,0,1,3),2,2,byrow = TRUE), widths=c(3,1), heights=c(1,3), TRUE)
layout.show(nf)
```


## Data visualization before `ggplot2` {.smaller}

### <span style="color:cornflowerblue">Figures with multiple panels:</span> the `layout` function

```{r, echo=F,eval=T, fig.height=4.5,fig.width=4.5}
nf <- layout(matrix(c(2,0,1,3),2,2,byrow = TRUE), c(3,1), c(1,3), TRUE)
#sample values from two normal distributions
x <- pmin(3, pmax(-3, stats::rnorm(50)))
y <- pmin(3, pmax(-3, stats::rnorm(50)))

xhist <- hist(x, breaks = seq(-3,3,0.5), plot = FALSE)
yhist <- hist(y, breaks = seq(-3,3,0.5), plot = FALSE)
top <- max(c(xhist$counts, yhist$counts))

xrange <- yrange <- c(-3, 3)
par(mar = c(3,3,1,1))
plot(x, y, xlim = xrange, ylim = yrange, xlab = "", ylab = "")
par(mar = c(0,3,1,1))
barplot(xhist$counts, axes = FALSE, ylim = c(0, top), space = 0)
par(mar = c(3,0,1,1))
barplot(yhist$counts, axes = FALSE, xlim = c(0, top), space = 0, horiz = TRUE)
```


## In-class exercise 2 {.smaller}


- Extend your exploration of `swiss` by using 2 or three of the figure types discussed after `plot`

- Make a figure with two panels (1 row by two columns) using `mfrow`

- Make a figure with 3 different plots using `layout`


<!-- ## Acknowledgements {.larger} -->

<!-- <span style="color:blue">Materials above are adapted from the following sources:</span> -->


<!-- - [Advanced R book](https://adv-r.hadley.nz/vectors-chap.html) -->
<!-- - <span style="color:forestgreen">Colin Rundel's notes from Duke University</span> -->
<!-- - [Kaplan et al, 2020](https://cran.r-project.org/web/packages/mosaic/vignettes/Resampling.pdf) -->
<!-- - [UW SISG Permutation Tests](http://faculty.washington.edu/kenrice/sisg/SISG-08-06.pdf) -->