---
title: 'Week 3: R Objects and Intro to Data Vis'
author: "Daniel Taylor-Rodriguez"
date: "April 14, 2020"
output:
  ioslides_presentation:
   incremental: true
subtitle: STAT 399
---

<style>
.forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width='\\textwidth')
```



## Plan for this Week {.build}

### <span style="color:cornflowerblue"> Part 1: wrap up base R </span>

1. Finish subsetting (matrices, factors and data.frames)

2. Subsetting and assignment

### <span style="color:cornflowerblue"> Part 2: modern R basics </span>

3. Visualization in base R 

4. Visualization in ggplot2 (intro)




# Subsetting

## Subsetting Matrices {.smaller .build}

```{r}
(x <- matrix(1:6, nrow=2, ncol=3))
```

<div class="columns-2">
```{r}
x[1,3]
x[1:2, 1:2]
```

<p class="forceBreak"></p>
```{r}
x[, 1:2]
x[-1,-3]
```
</div>


## Preserving Subsetting {.smaller .build}

Most of the time, R's `[` subset operator is a *preserving* operator, in that the returned object will have the same type as the parent. Confusingly, when used with a matrix or array `[` becomes a *simplifying* operator (does not preserve type) - this behavior is controlled by the `drop` argument.

<div class="columns-2">
```{r}
x[1, ]
x[1, , drop=TRUE]
x[1, , drop=FALSE]
```

<p class="forceBreak"></p>
```{r}
str(x[1, ])
str(x[1, , drop=TRUE])
str(x[1, , drop=FALSE])
```
</div>



## Factor Subsetting  {.smaller .build}

```{r}
(x <- factor(c("BS", "MS", "PhD", "MS")))
x[1:2]
x[1:2, drop=TRUE]
```


## Data Frame Subsetting  {.smaller .build}

If provided with a single value, data frames assume you want to subset a column or columns - multiple values then the data frame is treated as a matrix.

```{r}
df <- data.frame(a = 1:2, b = 3:4)
df[1]
df[[1]]
df[, "a"]
```


## Data Frame Subsetting  {.smaller .build}

<div class="columns-2">
```{r}
df["a"]
df[, "a", drop = FALSE]
```

<p class="forceBreak"></p>
```{r}
df[1,]
df[c("a","b","a")]

```
</div>

## Preserving vs Simplifying Subsets {.smaller .build}

Type             |  Simplifying             |  Preserving
:----------------|:-------------------------|:-----------------------------------------------------
Atomic Vector    |  `x[[1]]`                |  `x[1]`
List             |  `x[[1]]`                |  `x[1]`
Matrix / Array   |  `x[1, ]` <br/> `x[, 1]` |  `x[1, , drop=FALSE]` <br/> `x[, 1, drop=FALSE]`
Factor           |  `x[1:4, drop=TRUE]`   |  `x[1:4]`
Data frame       |  `x[, 1]` <br/> `x[[1]]` |  `x[, 1, drop=FALSE]` <br/> `x[1]`

# Subsetting and assignment 


## Subsetting and assignment {.smaller .build}

Subsets can also be used with assignment to update specific values within an object.

```{r}
x <- c(1, 4, 7)
```

```{r}
x[2] <- 2
x
x[x %% 2 != 0] = x[x %% 2 != 0] + 1
x
x[c(1,1)] = c(2,3)
x
```


## Subsetting and assignment {.smaller .build}

<div class="columns-2">
```{r}
x <- 1:6
x[c(2,NA)] <- 1
x
```


```{r}
x <- 1:6
x[c(TRUE,NA)] <- 1
x
```

<p class="forceBreak"></p>
```{r}
x <- 1:6
x[c(-1,-3)] <- 3
x
```

```{r}
x <- 1:6
x[] <- 6:1
x
```
</div>

## Deleting list (df) elements {.smaller .build}

```{r}
df <- data.frame(a = 1:2, b = TRUE, c = c("A", "B"))
```

```{r}
df[["b"]] <- NULL
str(df)
```

```{r}
df[,"c"] <- NULL
str(df)
```


## Subsets of Subsets {.smaller .build}

```{r}
df <- data.frame(a = c(5,1,NA,3))
```

```{r}
df$a[df$a == 5] <- 0
df
```

```{r}
df[1][df[1] == 3] <- 0
df
```

## Peer Review: Grading Rubric {.smaller}

<div style="color:cornflowerblue"> 
I will randomly assign graders to anonymized assigments, post a list of the file each of you needs to look at on D2L, and give you access on Rstudio Cloud to the files you are expected to look at. 
</div>

<span style="color:forestgreen">**Be constructive and considerate**</span>

0 - No homework turned in.


1 - Turned in but low effort, poorly presented with nonfunctional code and ignoring directions.


2 - Decent effort, well presented all code works and followed directions with some minor issues.


3 - Nailed it!

# In-class exercise

## Permutation test for two group means {.smaller .build}

In an experiment on memory (Mednicj et al, 2008), students were given lists of 24 words to memorize. After hearing the words they were assigned at random to different groups: 

- One group of 12 students took a nap for 1.5 hours while 
- A second group of 12 students stayed awake and was given a caffeine pill. 

These data contain the number of words each participant was able to recall after the break. 
```{r}
Memory <- read.csv("http://www.mosaic-web.org/go/datasets/SleepCaffeine.csv")
```

Go to the `week3_exe_Lastname_Firstname.Rmd` file in RStudio Cloud to work on the problem.








<!-- # Loading data into R -->

<!-- ## Loading data into R {.smaller .build} -->

<!-- From R we can access data stored in an online repository or simply load it from a file stored in our computer in MANY different formats.   -->

<!-- Some examples are `.txt`, `.csv` `.xlsx`, `.dat`, `.shp`, `.dbf`, etc.  -->

<!-- Today we'll only talk some of the most common of these.  We'll revisit other types later in the course. -->

<!-- ## Loading data from `url`'s {.smaller .build} -->



<!-- ## Loading `.txt` files {.smaller .build} -->

<!-- ### Loading `.txt` data separated by some character (could be a blank space: denoted by "") -->

<!-- 1. Use read.table(file= ,header= , sep= )  -->
<!-- ```{r} -->
<!-- # Use read.table(file= ,header= , sep= ) function to load the file "Education.txt" -->
<!-- # store it in an object called Edu.blank -->
<!-- Edu.blank <- read.table("Education.txt",header=T) -->
<!-- ``` -->









## Acknowledgements {.larger}

<span style="color:blue">Materials above are adapted from the following sources:</span>


- [Advanced R book](https://adv-r.hadley.nz/vectors-chap.html)
- <span style="color:forestgreen">Colin Rundel's notes from Duke University</span>
- [Kaplan et al, 2020](https://cran.r-project.org/web/packages/mosaic/vignettes/Resampling.pdf)
- [UW SISG Permutation Tests](http://faculty.washington.edu/kenrice/sisg/SISG-08-06.pdf)