---
title: "Homework 2"
author: "Do not include your name, keep anonymous for peer review"
date: 'Due: Wednesday 4/29/2020 by 12:00 pm'
output:
  prettydoc::html_pretty:
        theme: architect
        highlight: github
subtitle: 'STAT 399: Statistical Computing and Data Visualization with R'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggthemes)
```

## Instructions

- Make a copy of this template to write out your solution, and simply knit it as `HW02.Rmd`.
- Inside this .Rmd file do not include any personal identifier (such as your name, Odin ID, etc.). 
- Knit your Rmd file as html and upload both the Rmd and html files with your solution to D2L in `Activities > Assignments > Homework2` before Wednesday April 29th at 12:00 pm.


## Objectives for this week's homework

1. Practice making figures in both base R and ggplot2
2. Generate hypotheses and explore them graphically


Write down a question or hypothesis you are interested in exploring from the *world powerplant dataset* I found in the [information is beautiful](https://informationisbeautiful.net/visualizations/mega-what-the-worlds-biggest-and-most-notable-power-plants/) webpage, which includes some nice visualizations.  The original data can be found [here](http://datasets.wri.org/dataset/globalpowerplantdatabase) along with a brief description of the data.  Note that inside of the globalpowerplant folder in RStudio Cloud (as well as in the D2L the homework folder) I provide additional documents with information that may prove useful, in particular be sure to look at the `README.txt` file, which includes the metadata (a description of the variables) in the **Fields** section.  

The idea is for you to investigate your question/hypothesis graphically.  To do so, generate the figures you consider necessary -- at least two and at most 4, and note that more figures is not necessarily better, fewer but very well though-out and informative figures might do the trick. The only condition is that <span style="color:orange">at least one of your figures is built using `base R` and at least one is built in `ggplot2`.*</span>

You might want to work with just a subset of the data (e.g. one particular country), or you may choose to transform or summarize your data in the way that best allows you to explore your question. All titles, axes, and legends should be labelled clearly (no raw variable names). You must have at least one graph with `facet_wrap` or `facet_grid`. You must include at least one manually specified legend. You can use other `geoms` like histograms, bar charts, add vertical or horizontal lines, etc. [You may find this data visualization cheat sheet helpful](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf).

Your document should be aesthetically pleasing, with some organization. You must write up your observations in words as well as showing the graphs. Use chunk options `echo` and `results` to limit the code/output you show in the `.html`.

Below I load n the data, but be careful, you may need to filter out rows that have `NA`s in some of specific variables, so recall that you can test (and filter out) for `NAs` using the function `is.na`, which is vectorized (look at week 2's classnotes)!

Have fun and Good Luck!!
```{r load_data}
power <- read.csv('globalpowerplant/global_power_plant_database.csv',
                  na.strings = c(NA,"","NA"),
                  header=T)
str(power)
```

# Global Power

```{r libraries, echo=F}
library(ggplot2)
library(dplyr)
require(maps)
library(repr)
```

```{r peak function}
# takes a look at the first n/2 and last n/2 rows. If the number is
# odd then the extra row is given to the tail
peak <- function( df, rows = 10L ){
  hrow = as.integer(rows/2L)
  trow = rows - hrow
  return(rbind(head(df,hrow), tail(df,trow)))
}
```

```{r}
peak(power,20)
unique(power$primary_fuel)
```

Let's look at a splattering of location, source, and capacity. We'll use the `maps` package to be able to draw the map, then use the gps locations
to place the points. Each point will be a bubble size of generation capacity, and the color of the bubble will be the primary fuel source.


```{r world map}
world_map <- map_data("world")
ggplot(world_map, aes(x=long, y=lat, group=group)) +
  geom_polygon(fill="lightgray", colour="white") +
  xlab("") + ylab("") +
  theme(legend.position = "none")
```


We want to merge the map data and the data for power plants. To do this we'll run some selection on the `power` dataframe and rename the `country_long` column to the `country` column and follow that by a left outer join.

First we'll take care of cleaning up the generation. Some rows have NA across the board, these we will want to preserve the NAness of it to remove them from further
processing, but for the ones that have at least one value in the years we'll or the estimated amount, we'll want to calculate a column that uses the available
information.

To do this we'll make a function, then apply it as a new column to a new dataframe.

```{r}
estimated_power <- function( x ){
  if(!is.na(x["estimated_generation_gwh"])){
    return(x["estimated_generation_gwh"])
  }
  
  cols <- paste("generation_gwh",2013:2017,sep="_")
  
  generation <- unlist(x[cols])
  if(sum(!is.na(generation))){
    return(mean(generation,na.rm=TRUE))
  }else{
    return(NA_real_)
  }
}
```

Now let's test this out first on a subset of the data

```{r}
# Appraently `apply` coerces each row into a vector rather than a list
# This then causes a bunch of headaches with it returning a character type vector
# whereas I want a numeric type
# by passing in only the numeric columns I need for the function it works without a lot of
# double checking in the function.
cols <- c( "estimated_generation_gwh", paste("generation_gwh",2013:2017,sep="_"))
result <- apply( power[sample(seq_along(1:nrow(power)), size=10 ), cols], 1, FUN=estimated_power )

str(result)
```

We've tested this out on a random sample, now we'll try this out on the full set. We'll also make
some splits on the data for generation as there are some plants that make less energy than they
expend, some are neutral, and most create more energy than they expend.

```{r}
powermap <- cbind( power[,c("country_long","longitude", "latitude","primary_fuel")], mean_generation_gwh=apply(power[,cols], 1, FUN=estimated_power ) )
names(powermap)[names(powermap) == "country_long"] <- "region"

pmap.power <- na.omit(powermap)
pmap.positive.power <- pmap.power[which(pmap.power$mean_generation_gwh > 0),]
pmap.negative.power <- pmap.power[which(pmap.power$mean_generation_gwh < 0),]
pmap.neutral.power <- pmap.power[which(pmap.power$mean_generation_gwh ==0),]
pmap.world <- world_map[,c("long", "lat", "group", "order", "region")]

```

Since we're going to be making a few maps that are the same but for the data they work on
we'll make a function factory to do this for us. The `geom_point` will be the most likely
candidate to work with.

```{r}
plot_data <- function( data, map_plot ){
  ggplot() +
  geom_polygon( data=map_plot, aes(x=long, y=lat, group=group), fill="lightgray" ) +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(),
        axis.ticks.x = element_blank(), axis.text.x = element_blank() ) +
  xlab("") + ylab("") +
  geom_point(data=data,aes(x=longitude, y=latitude, color = primary_fuel, size = mean_generation_gwh), alpha=0.25, ) +
    labs(color="Primary Fuel") +
  theme(legend.position = "top")
}
```


```{r}
pmap.negative.map <- plot_data(pmap.negative.power, pmap.world) +
  scale_size_continuous(range = c(.1, 10), trans = "reverse", guide=(All=FALSE)) +
  ggtitle("Negative Power Plants") 
pmap.negative.map
```

Seems that the only place in the world with negative power plants are in the US and maybe canada.


```{r}
pmap.map <- plot_data(transform(pmap.power, fct = cut(mean_generation_gwh, c(-5e+04,-1e-01,1e-01,5e+07))), pmap.world)
pmap.primary_fuel_facet <- pmap.map +
  scale_size_continuous(range = c(.1, 10), trans = "identity", guide=(All= FALSE) ) +
  facet_wrap(~fct)

pmap.primary_fuel_facet
```

This is looking nice, but it's a bit clutered and hard to see. Unfortunately this doesn't have a column called "Continent" or else we would
create some facets based on continent to hopefully have a better view. 
Instead we'll make our own grouping of energy categories and do a
left-join onto it

```{r}

unique(pmap.power$primary_fuel)
```

Here we see the 15 factors used. We'll categorize them ourselves and then join them in.

```{r}
pf = data.frame(primary_fuel=unique(pmap.power$primary_fuel))
pf = cbind(pf, fuel_category=c("Green", "Unknown", "Black", "Black", "Green", "Black", 
                               "Green", "Green", "Unknown", "Green", "Black", "Green", 
                               "Green", "Unknown"))

```


Now we'll left join it

```{r}
pmap.category_power <- left_join(pmap.power, pf, by=c("primary_fuel"))
```


```{r}
pmap.map <- plot_data(pmap.category_power, pmap.world)
pmap.primary_fuel_category_facet <- pmap.map +
  scale_size_continuous(range = c(.1, 10), trans = "identity", guide=(All= FALSE) ) +
  facet_wrap(~fuel_category) +
  options(repr.plot.width=800, repr.plot.height=500)

pmap.primary_fuel_category_facet
```


Now we're sort of getting somewhere, we see a large distribution of black and green energy
throughout the globe. Even though it's an interesting looking plot, it doesn't really tell
us much.

Let's narrow the plot

```{r fig.width=7, fig.height=12}
pmap.usa <- plot_data(pmap.category_power[which(pmap.category_power$region=="United States of America" & pmap.category_power$longitude < 144),], pmap.world[which(pmap.world$region=="USA" & pmap.world$long < 144),])
pmap.usa_primary_fuel_category_facet <- pmap.usa +
  scale_size_continuous(range = c(.1, 10), trans = "identity", guide=(All= FALSE) ) +
  facet_wrap(~fuel_category, ncol=1) +
  ggtitle("US Energy Generation by Non/Renewable Sources")
  
pmap.usa_primary_fuel_category_facet
```

This one shows us visually the concentration of power generation by fuel type category. Here we see
a green suspender going up the West Coast. We also notice there are some areas where black energy
sources are not present but green sources are, especially in the mid-west.

## Base R

Let's try to look at the annual generation by year. We'll take a slice of the data and
turn it into a more appropriate

```{r}
baser.cols <- c(paste("generation_gwh", 2013:2017, sep="_" ))
baser.power <- power[,baser.cols]
baser.byyear <- apply(baser.power, 2, sum, na.rm=TRUE)
baser.df <- data.frame(Year=c(2013:2017))
baser.df <- cbind(baser.df, Generation=baser.byyear )


barplot(Generation ~ Year, data=baser.df,
        ylim=c(9e+5,6e+6),
        xlab = "Year",
        ylab = "Generation")
title("World Energy Generation by Year")
```


