---
title: "Exercises Week 6"
author: "David Post"
date: "5/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(readr)
library(tidyr)
library(dplyr)
library(purrr)
library(ggplot2)
library(tidyverse)
library(nycflights13)
library(gapminder)
```

## Tuesday Lecture

### Problem 1

Compute the rate for `table2`, and `table4a + table4b`. You will need to perform four operations:

1. Extract the number of TB cases per country per year
2. Extract the matching population per country per year
3. Divide cases by population, and multiply by 10000
4. Store back in the appropriate place.

```{r message=F,warning=F}
library(tidyverse)
table2
cases <- table2 %>% .[.$type=="cases",] %>% pull(count) 
table2 %>% 
  .[.$type=="population",] %>% 
  mutate(rate=(cases/count)*10000) %>% 
  select(-type,-count)
```


```{r eval=F,message=F,warning=F}
table4a
combined.4a <- table4a %>%
  rename(cases_1999="1999", cases_2000="2000")

combined.4b <- table4b %>%
  rename(pop_1999="1999", pop_2000="2000")

combined.4ab <- merge(combined.4a,combined.4b,by=c("country"))

combined.4ab %>%
  mutate(rate_1999=(.$cases_1999/.$pop_1999)*10000) %>% 
  mutate(rate_2000=(.$cases_2000/.$pop_2000)*10000)
#table4a %>% gather(key=year,value=counts,-country) %>% pull()
```

### Problem 2

Use tb to generate a tibble having as columns iso2, gender and age, as well as the counts for each year from 1980 through 2008.

```{r}
tb = readr::read_csv(paste0("https://github.com/tidyverse/",
                            "tidyr/raw/master/vignettes/tb.csv"))

tb %>% 
  gather(key=group, value=counts, 
         -iso2, -year, na.rm = TRUE) %>%
  separate(col=group, into = c("gender","age"), sep=1) %>%
  mutate(
    age = case_when(
      age == "04" ~ "0-4",     age == "514" ~ "5-14",
      age == "014" ~ "0-14",   age == "1524" ~ "15-24",
      age == "2534" ~ "25-34", age == "3544" ~ "35-44",
      age == "4554" ~ "45-54", age == "5564" ~ "55-64",
      age == "65" ~ "65+",     age == "u" ~ NA_character_
    )
  ) %>%
  spread(key=year, value=counts)
```

