---
title: "Week1_exe_Post_D"
author: "David Post"
date: "4/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set 1 -- Exercise 1  {.build .smaller}

Write a set of conditional(s) that satisfies the following requirements:

* If `x` is greater than 3 and `y` is less than or equal to 3 then print "Hello world!"

* Otherwise if `x` is greater than 3 print "!dlrow olleH"

* If `x` is less than or equal to 3 then print "Something else ..."

* Stop execution if x is odd and y is even and report an error, don't print any of the text strings above.

Test out your code with

1. `x=1` and `y=2`
2. `x=1` and `y=3` 
3. `x=3` and `y=3`

```{r}
check_conditions <- function( x, y ){

  if( (x %% 2) && !(y %% 2) )
      stop("Error, x is odd and y is even")

  if( ( x > 3 ) && ( y <= 3 ) ) {
      print("Hello World!")
  }else if ( x > 3 ){
    print("!dlroW olleH")
  }
  
  if( x <= 3 )
    print("Something else...")
}
data1 <- vector( mode="list", length=3)
data1[[1]] <- c(1,2)
data1[[2]] <- c(1,3)
data1[[3]] <- c(3,3)

# This doesn't work because the first dataset
# runs into the stop clause and ends any subsequent
# execution

#for( i in data1){
#  check_conditions(i[[1]],i[[2]])
#}
```

```{r, error=TRUE}
#check_conditions(data1[[1]][[1]],data1[[1]][[2]])
```
```{r}
check_conditions(data1[[2]][[1]],data1[[2]][[2]])
```
```{r}
check_conditions(data1[[3]][[1]],data1[[3]][[2]])
```

## Set 1 -- Exercise 2 {.build .smaller}

Write a set of conditional(s) in `R` to solve equations of the form $$a x^2 + b x + c,$$ which accounts for possible errors that might occurr in the calculation (you may use `if`, `else if`, `else`, `stop`, `stopifnot`)

Test out your code with

1. `a=1`, `b=0` and `c=1`
2. `a=1`, `b=0` and `c=-1`
3. `a=9`, `b=4` and `c=1`
4. `a=0`, `b=3` and `c=3`
4. `a=1`, `b=-4` and `c=4`



```{r, error=TRUE}
quadratic <- function(a,b,c){
  deter <- b*b-4*a*c
  stopifnot( b*b-4*a*c > 0 )
  if( deter < 0 )
    return(-1)
  
  # Update the determinant to be the squareroot of
  # the determinant. If we've made it this far then
  # we know we stay within our field.
  deter <- sqrt(deter)
  first <- (-b + deter)/2 
  second <- (-b - deter)/2
  if( first == second ){
    cat("Single root: ", first, "\n")
  }else{
    cat("First root: ", first, "\n")
    cat("Second root:", second, "\n")
  }
}


quadraticv <- function(v){
  quadratic(v[[1]], v[[2]], v[[3]])
}
  
data <- vector(mode="list", length=5)
data[[1]] <- c( 1,0,1)
data[[2]] <- c( 1,0,-1)
data[[3]] <- c(9,4,1)
data[[4]] <- c(0,3,3)
data[[5]] <- c(1,-4,4)

# Because of the stop statement this for loop
# will not work for our intended purposes.

#for( i in data ){
#  quadraticv(i)
#}

```

```{r, error=TRUE}
  #quadraticv(data[[1]])
```

```{r}
  quadraticv(data[[2]])
```

```{r, error=TRUE}
  #quadraticv(data[[3]])
```

```{r}
  quadraticv(data[[4]])
```

```{r, error=TRUE}
  #quadraticv(data[[5]])
```

## Set 2 -- Exercise  1 {.smaller}

Below is a vector containing all prime numbers between 2 and 100:

<div class="centered">
```r
primes = c( 2,  3,  5,  7, 11, 13, 17, 19, 23, 29, 31, 37, 41, 
      43, 47, 53, 59, 61, 67, 71, 73, 79, 83, 89, 97)
```
</div>

If you were given the vector 

<div class="centered">
`x = c(3,4,12,19,23,51,61,63,78)`, 
</div>
write the R code necessary to print only the values of `x` that are *not* prime (without using subsetting or the `%in%` operator). 

Your code should use *nested* loops to iterate through the vector of primes and `x`.
```{r}
primes = c( 2,  3,  5,  7, 11, 13, 17, 19, 23, 29, 31, 37, 41, 
      43, 47, 53, 59, 61, 67, 71, 73, 79, 83, 89, 97)

x = c(3,4,12,19,23,51,61,63,78)


for( i in x ){
  isprime <- TRUE # Reset sentinal
  for( prime in primes ){
    if( i == prime ){
      isprime <- FALSE # Found a prime
      break; # Short circuit out of inner loop
    }
  }
  
  if( isprime )
    print( i )
}

```



