---
title: 'Week 5: Data Visualization and Data Manipulation'
author: "Daniel Taylor-Rodriguez"
date: "April 30, 2020"
output:
  ioslides_presentation:
   incremental: false
subtitle: STAT 399
---

<style>
.forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>

```{r, message=FALSE, warning=FALSE, include=FALSE}
options(
  htmltools.dir.version = FALSE, # for blogdown
  width = 80,
  tibble.width = 80
)

knitr::opts_chunk$set(
  fig.align = "center"
)

htmltools::tagList(rmarkdown::html_dependency_font_awesome())

library(tidyr)
library(dplyr)
library(purrr)
library(magrittr)
library(ggplot2)
library(shiny)
library(ggthemes)
library(nycflights13)
```



## Plan for this Week {.build}

1. Wrapping up [`ggplot2`](https://r4ds.had.co.nz/data-visualisation.html) basics: *statistical transformations*

2. `shinyr` teaser

3. Data and `tibbles`

4. Pipes in `magritt`

5. Data manipulation with `dplyr`



## Loading the data (from a url) {.smaller}

```{r message=FALSE, echo=TRUE} 
# Before we were uploading the data from the file locally, using:
# movies = readr::read_csv("data/movies.csv"), now we are pulling the file 
# directly from github. To do so, you need the url to the "raw" version
# of the file
moviesdatapath <- "https://raw.githubusercontent.com/mine-cetinkaya-rundel/df2016_workshops/master/viz_ggplot2_shiny/data/movies.csv"
movies = readr::read_csv(url(moviesdatapath))
movies
```



# Statisical Transformations

## Statisical Transformations

```{r eval = F}
geom_bar(mapping = NULL, 
         data = NULL, 
         stat = "bin", 
         position = "stack",...)
```

- `stat` statistically transforms input data (bin means bin and count)

- `position` dodges for side-by-side bars or stack for additive bars


## Statisical Transformations

```{r fig.height=3.5, fig.width=5, warning=F, message=F}
ggplot(data = movies, aes(x = audience_score)) + geom_bar(stat="bin") 
```


## Statisical Transformations

```{r fig.height=3.5, fig.width=5, warning=F, message=F}
ggplot(data = movies, aes(x = audience_score)) + 
  geom_bar(stat="bin", binwidth = 20) 
```

`binwidth` specifies the number of bins


## Let's do an exercise together {.smaller .build}

- Do the transformation stat="bin" by hand

- Cut `audience_score` into groups (0,20] (20,40] (40,50] (60,80] (80,100] with 
```{r eval=F}
cut(movies$audience_score, breaks=seq(0,100,5))
```

- Count observations by group and save with
```{r eval=F}
count.df <- as.data.frame(table(movies$audience_score))
```

- Use ggplot() + geom_bar()

**Question: what `stat` argument do you need to make the barplot with `count.df`?**

## Let's do an exercise together {.smaller .build}

```{r}
aud_scorecut <- cut(movies$audience_score, breaks=seq(0,100,20))
count.df <- as.data.frame(table(aud_scorecut))
count.df
```
## Let's do an exercise together {.smaller .builder}

```{r eval=F}
ggplot(count.df, aes(x=aud_scorecut, y=Freq)) + 
  geom_bar(stat="bin")  
```

<span style="color:forestgreen"> No good, why? </span>

```{r echo=F, error=T}
ggplot(count.df, aes(x=aud_scorecut, y=Freq)) + 
  geom_bar(stat="bin")  
```

## Let's do an exercise together {.smaller .build}

Ok, so how about this? 
```{r error=T, eval=T}
ggplot(count.df, aes(x=aud_scorecut)) + 
  geom_bar(stat="bin")  
```


## Let's do an exercise together {.smaller}

```{r fig.height=3.5, fig.width=5}
ggplot(count.df, aes(x=aud_scorecut, y=Freq)) + 
  geom_bar(stat="identity")  
```




# Plug for possible extra topic: Shiny

## Shiny

- A framework for R to easily turn your analyses into interactive web apps

- No HTML, CSS, or JavaScript knowledge required


## Live demo

```{r echo = FALSE, eval = FALSE}
# Define UI for application that draws a ggplot
ui <- shinyUI(fluidPage(
  
  # Application title
  titlePanel("Exploring Movies"),
  
  # Sidebar with a slider input for number of bins 
  sidebarLayout(
    sidebarPanel(
      # Select variable for y-axis
      selectInput("y",
                  "Y-axis:",
                  choices = c("imdb_rating", "imdb_num_votes",
                              "critics_score", "audience_score",
                              "runtime"),
                  selected = "audience_score"
      ),
      
      # Select variable for x-axis
      selectInput("x",
                  "X-axis:",
                  choices = c("imdb_rating", "imdb_num_votes",
                              "critics_score", "audience_score",
                              "runtime"),
                  selected = "critics_score"
      ),
      
      # Select variable for color
      selectInput("z",
                  "Color by:",
                  choices = c("title_type", "genre", "mpaa_rating",
                              "critics_rating", "audience_rating")
      ),
      
      # Alpha
      sliderInput("alpha", 
                  "Alpha:", 
                  min = 0, max = 1, 
                  value = 0.5),
      
      # Horizontal line separating population and sample input
      hr(),
      
      # Sample size
      numericInput("n",
                   "Sample size:",
                   min = 1,
                   max = nrow(movies),
                   value = nrow(movies))
      
    ),
    
    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("exp_plot")
    )
  )
))

# Define server logic required to draw a histogram
server <- shinyServer(function(input, output) {
  
  # Create a new data frame that is a sample of n observations from movies
  movies_sample <- reactive({
    movies %>%
      sample_n(input$n)
  })
  
  output$exp_plot <- renderPlot({
    ggplot(data = movies_sample(), aes_string(x = input$x, y = input$y, 
                                              color = input$z)) +
      geom_smooth(method=lm) +
      geom_point(alpha = input$alpha)
  })
})

# Run the application 
shinyApp(ui = ui, server = server)
```

# Tibbles


## Modern data frames

`tibble` is a package that *modernizes* data frames

```{r}
library(tibble)
class(iris)
tbl_iris <- as_tibble(iris)
class(tbl_iris)
```

## Tibbles vs data.frames {.smaller}


### <span style="color:cornflowerblue">Two main differences</span>

* <span style="color:forestgreen">Printing:</span> Print method shows only the first 10 rows and all columns that fit on screen.  Also, the type of each column is reported

* <span style="color:forestgreen">Subsetting:</span> 
 * `[` with tibble always returns a tibble, whereas with `data.frame` some times return vectors and in others data frames.
 * Tibbles never do partial matching, and they will generate a warning if the column you are trying to access does not exist

## Fancy Printing

```{r}
tbl_iris
```




## Subsetting: tibbles are lazy {.smaller}

```{r}
tbl_iris[1,]
```

<div class="columns-2">

```{r}
tbl_iris[,"Species"]
```

<p class="forceBreak"></p>
```{r}
tibble(
  x = 1:3, 
  y = c("A","B","C")
)
```
</div>


## Reverting a tbl {.smaller}

```{r}
d <- tibble(
  x = 1:3, 
  y = c("A","B","C")
)

d
```


<div class="columns-2">
```{r}
data.frame(d)
```


<p class="forceBreak"></p>
```{r}
class(d) <- "data.frame"
d
```
</div>

## Multiple classes {.smaller}

```{r}
d <- tibble(
  x = 1:3, 
  y = c("A","B","C")
)

class(d)
```


```{r}
class(d) <- rev(class(d))
class(d)
d
```

# Nesting Functions with Pipes



## Pipes in R {.smaller .build}

Think about the following sequence of actions 

<span style="color:cornflowerblue"> *find key $\Rightarrow$ unlock car $\Rightarrow$ start car $\Rightarrow$ drive to school $\Rightarrow$ park* </span>

- In R pseudocode, these as nested functions could look like:
```{r eval=FALSE}
park(drive(start_car(find("keys")), to="campus"))
```

- [Pipes](https://r4ds.had.co.nz/pipes.html) provide a better flowing and more readable structure:

```{r eval=FALSE}
find("keys") %>%
    start_car() %>%
    drive(to="campus") %>%
    park()
```


## Approaches {.smaller}

These all work, it just comes down to personal preference

Nested:
```{r, eval=FALSE}
h( g( f(x), y=1), z=1 )
```

Piped:
```{r, eval=FALSE}
f(x) %>% g(y=1) %>% h(z=1)
```

Intermediate:
```{r, eval=FALSE}
res <- f(x)
res <- g(res, y=1)
res <- h(res, z=1)
```


## What about other arguments?  {.smaller .build}

<span style="color:cornflowerblue">If we want to</span>

1. send results to a function argument other than first one or 
2. use the previous result for multiple arguments.

<span style="color:cornflowerblue">we can refer to the result from previous pipe with `.`</span>

<div class="columns-2">

```{r}
data.frame(a=1:3,b=3:1) %>% 
  lm(a~b,data=.)
```

<p class="forceBreak"></p>
```{r}
data.frame(a=1:3,b=3:1) %>% .[[1]]
```

```{r}
data.frame(a=1:3,b=3:1) %>% 
  .[[length(.)]]
```
</div>

## In-class example  {.smaller .build}

<div class="centered">
<div style="color:green">

*Little bunny Foo Foo*

*Went hopping through the forest*

*Scooping up the field mice*

*And bopping them on the head*
</div>
</div>

1. Start with the object `FooFoo <- little_bunny()`, and consider the verbs (i.e., functions)  `hop()`, `scoop()`, `bop()`.

2. Using `FooFoo` and the verbs, retell the story in code in all four ways (defining additional inputs as needed):

* Saving each intermediate step as a new object
* Overwriting the original object many times
* Composing functions
* Using pipes


## Should you always use pipes? {.smaller}

### <span style="color:orange">NO!! They are pretty useful but do not solve every problem! </span>

Pipes most useful for writing shorter linear sequences of operations. 

<span style="color:orange">**Consider using other options with:**</span>

1. very long sequence of operations

2. multiple inputs or outputs

3. problems with complex dependencies

4. functions that use the current environment

5. functions that use lazy evaluation


## Recap on pipes {.smaller .build}

<span style="color:cornflowerblue">Objects on the left of `%>%` passed to first argument of function on the right:</span>

* The piped expression `x %>% g()` yields `g(x)`

<span style="color:cornflowerblue">Other arguments go on the right in the function:</span>

* The piped expression `x %>% f(y)` yields `f(x,y)`

<span style="color:cornflowerblue">If you pipe a function where data not first argument, use `.` in data argument</span>

* The piped expression `x %>% h(y,.)` yields `h(y,x)`

### <span style="color:forestgreen">Keybord shortcut for `%>%` is </span>

* `Ctrl+Shift+M` or `Command+Shift+M`


## In-class Exercise {.smaller .build}

Using the `data.frame` `X` below, we want to:

* Create a crosstabulated table for the first and second `x` and `y` (use `table` or `xtabs`)

* Count the number of cases for each value of `y` (use `table`)

* Find `mode` among levels of `x` (use `table` combined with `max` and `names`)

```{r eval=T}
X <- data.frame(x=sample(c("q","l","m"),50,replace=T),
                y=sample(c("a","b","c"),50,replace=T),
                z=sample(c("c","d","e"),50,replace=T))
str(X)
```

## In-class Exercise

###  <span style="color:forestgreen">Do this in three ways:</span>

* Taking one intermediate step at a time save to same object
* Using composite functions
* Using pipes



# Manipulating Data: dplyr

## A Grammar of Data Manipulation {.smaller .build}

`dplyr` based on concepts of functions as verbs manipulating data frames.

Single data frame functions / verbs:

* `filter()`: pick rows matching criteria
* `slice()`: pick rows using index(es)
* `select()`: pick columns by name
* `pull()`: grab a column as a vector
* `rename()`: rename specific columns
* `arrange()`: reorder rows
* `mutate()`: add new variables
* `transmute()`: create new data frame with variables
* `distinct()`: filter for unique rows
* `sample_n()` / `sample_frac()`: randomly sample rows
* `summarise()`: reduce variables to values
* ... (many more)


## dplyr rules for functions

<div style="color:forestgreen">
1. First argument is *always* a data frame

2. Subsequent arguments say what to do with that data frame

3. *Always* return a data frame

4. Don't modify in place

</div>



## A Working Example {.smaller}

Let's see `dplyr` in action with the nycflights13 data.

```{r message=FALSE}
flights
```



## <span style="color:cornflowerblue">`filter()` - March flights</span> {.smaller}

```{r}
flights %>% 
  filter(month == 3) #%>% head(4)
```



## <span style="color:cornflowerblue">`filter()` - Flights first 7 days of March </span>  {.smaller}

```{r}
flights %>% 
  filter(month == 3, day <= 7)
```



## <span style="color:cornflowerblue">`filter()` - Flights to LAX *or* RDU in March</span>  {.smaller}

```{r}
flights %>% 
  filter(dest == "LAX" | dest == "RDU", month==3)
```

## <span style="color:cornflowerblue">`filter()` - NY to PDX in April and June 1st</span>   {.smaller}

### <span style="color:cornflowerblue">Using `%in%`</span>
```{r}
flights %>% 
  filter(dest == "PDX", month%in%c(4,6), day==1)
```


## <span style="color:cornflowerblue">`slice()` - First 10 flights</span>  {.smaller}

```{r}
flights %>% slice(1:10)
```



## <span style="color:cornflowerblue">`slice()` - Last 5 flights</span>  {.smaller}

```{r}
flights %>% slice((n()-4):n()) 
```



## <span style="color:cornflowerblue">`select()` - Specific Columns</span>  {.smaller}

```{r}
flights %>% select(year, month, day) %>% head(4)
```



## <span style="color:cornflowerblue">`select()` - Exclude Columns</span>  {.smaller}

```{r}
flights %>% select(-year, -month, -day) %>% head(4)
```


## <span style="color:cornflowerblue">`select()` - Ranges</span>  {.smaller}

```{r}
flights %>% select(year:day) %>% head(4)
```



## <span style="color:cornflowerblue">`select()` - Exclusion Ranges</span>  {.smaller}

```{r}
flights %>% select(-(year:day)) %>% head(4)
```


## <span style="color:cornflowerblue">`select()` - Matching</span>  {.smaller}

```{r}
flights %>% select(contains("dep"), 
                   contains("arr")) %>% head(4)
```

## <span style="color:cornflowerblue">`select()` - Matching</span>  {.smaller}

```{r}
flights %>% select(starts_with("dep"), 
                   ends_with("time")) %>% head(4)
```

<span style="color:forestgreen">Other helpers: `ends_with`, `matches`, `num_range`, `one_of`, `everything`, `last_col`.</span>



## <span style="color:cornflowerblue">`select_if()` - Get non-numeric columns</span>  {.smaller}

```{r}
flights %>% select_if(function(x) !is.numeric(x))
```



## <span style="color:cornflowerblue">`pull()`</span>  {.smaller}

```{r}
names(flights)

# flights %>% pull("year") %>% head()
# 
# flights %>% pull(1) %>% head()

flights %>% pull(-1) %>% head()
```




## <span style="color:cornflowerblue">`rename()` - Change column names</span>  {.smaller}

```{r}
flights %>% 
  rename(tail_number = tailnum)
```
 


## <span style="color:cornflowerblue">`select()` vs. `rename()`</span>  {.smaller}

```{r}
flights %>% 
  select(tail_number = tailnum)
```

## <span style="color:cornflowerblue">`select()` vs. `rename()`</span>  {.smaller}

```{r}
flights %>% 
  rename(tail_number = tailnum)
```


## <span style="color:cornflowerblue">`arrange()` - Sort data </span> {.smaller}

```{r}
flights %>% 
  filter(month==3,day==2) %>% 
  arrange(origin, dest)
```



## <span style="color:cornflowerblue">`arrange()` & `desc()` - Descending order </span> {.smaller}

```{r}
flights %>% 
  filter(month==3,day==2) %>% 
  arrange(desc(origin), dest) %>% 
  select(origin, dest, tailnum)
```



## <span style="color:cornflowerblue">`mutate()` - Modify columns</span>  {.smaller}

```{r message=FALSE}
flights %>% 
  select(year:day) %>% 
  mutate(date = paste(year,month,day,sep="/"))
```



## <span style="color:cornflowerblue">`transmute()` - Create new tibble from existing columns</span>  {.smaller}

```{r}
flights %>% select(year:day) %>% transmute(date = paste(year,month,day,sep="/"))
```

## <span style="color:cornflowerblue">`transmute()` and `ifelse`</span>  {.smaller}

```{r}
flights %>% 
  transmute(arrive = ifelse( test = (arr_delay>0), 
                             yes = "late", no = "early" ) ) %>% head(6)
```



## <span style="color:cornflowerblue">`distinct()` - Find unique rows</span>  {.smaller}

```{r}
flights %>% select(origin, dest) %>% distinct() %>% arrange(origin,dest)
```



## <span style="color:cornflowerblue">Sampling rows </span> {.smaller}

```{r}
flights %>% sample_n(10)
```

## <span style="color:cornflowerblue">Sampling rows </span> {.smaller}

```{r}
flights %>% sample_frac(0.00003)
```



## <span style="color:cornflowerblue">`summarise()` </span> {.smaller}

```{r}
flights %>% summarize(n(), min(dep_delay), max(dep_delay))
```



```{r}
flights %>% 
  summarize(
    n = n(), 
    min_dep_delay = min(dep_delay, na.rm = TRUE), 
    max_dep_delay = max(dep_delay, na.rm = TRUE)
  )
```



## <span style="color:cornflowerblue">`group_by()` </span> {.smaller}

```{r}
flights %>% group_by(origin)
```
<span style="color:forestgreen">Huh? is this doing anything?</span>


## <span style="color:cornflowerblue">`group_by()` </span> {.smaller}

```{r}
flights %>% group_by(origin)
```
<span style="color:forestgreen">Prepares a *grouped table* to be used by other methods for aggregating by groups</span>

## <span style="color:cornflowerblue">`group_by()` </span> {.smaller}

```{r}
flights %>% group_by(origin,carrier)
```
<span style="color:forestgreen">Only apparent difference is that the grouping variable and the number of groups are specified</span>

## <span style="color:cornflowerblue">`summarise()` with `group_by()` </span> {.smaller}

```{r}
flights %>% group_by(origin) %>%
  summarize(
    n = n(), 
    min_dep_delay = min(dep_delay, na.rm = TRUE), 
    max_dep_delay = max(dep_delay, na.rm = TRUE)
  )
```

## <span style="color:cornflowerblue">`summarise()` with `group_by()` </span> {.smaller}

```{r}
flights %>% group_by(origin, carrier) %>%
  summarize(
    n = n(),
    min_dep_delay = min(dep_delay, na.rm = TRUE),
    max_dep_delay = max(dep_delay, na.rm = TRUE)
  ) %>%
  filter(n > 10000)
```


## <span style="color:cornflowerblue">`mutate()` with `group_by()` </span> {.smaller}

```{r}
flights %>% group_by(origin) %>%
  mutate(
    n = n(), 
  ) %>%
  select(origin, n)
```

## <span style="color:cornflowerblue"> `filter()`, `transmute()`, `ifelse`, `group_by()` and `summarise()`</span>  {.smaller}

```{r}
flights %>% filter(!is.na(arr_delay),) %>% 
  transmute(arr_delay_cat = ifelse( test = (arr_delay>0), 
                                    yes = "late", no = "early" ) ) %>% 
  group_by(arr_delay_cat) %>% 
  summarise(n=n())
```

## <span style="color:cornflowerblue"> Same but recoding with `recode`</span>  {.smaller}

```{r}
flights %>% transmute(arr_delay_cat = ifelse( test = (arr_delay>0), 
                                    yes = "late", no = "early" ) ) %>% 
  group_by(arr_delay_cat) %>% 
  summarise(n=n()) %>% 
  mutate(arr_delay_cat=recode(arr_delay_cat,late="bad",early="good"))
```
## <span style="color:cornflowerblue"> Same but recoding `NA` with `coalesce`</span>  {.smaller}

```{r}
flights %>% transmute(arr_delay_cat = ifelse( test = (arr_delay>0), 
                                    yes = "late", no = "early" ) ) %>% 
  group_by(arr_delay_cat) %>% 
  summarise(n=n()) %>% 
  mutate(arr_delay_cat=coalesce(arr_delay_cat,"no info"))
```

## <span style="color:cornflowerblue"> Recoding and changing `NA`'s single step</span>  {.smaller}

```{r}
flights %>% transmute(arr_delay_cat = ifelse( test = (arr_delay>0), 
                                    yes = "late", no = "early" ) ) %>% 
  group_by(arr_delay_cat) %>% 
  summarise(n=n()) %>% 
  mutate(arr_delay_cat=recode(arr_delay_cat,
                              late="bad",
                              early="good",
                              .missing="no info"))
```

## In-class Exercises  {.smaller .build}

1. How many flights to Los Angeles (LAX) did each of the legacy carriers (AA, UA, DL or US) have in May from JFK, and what was their average duration?

2. What was the shortest flight out of each airport in terms of distance? In terms of duration?

3. Which plane (check the tail number) flew out of each New York airport the most?

<div style="color:forestgreen">
*Optional*

Which date should you fly on if you want to have the lowest possible average departure delay? What about arrival delay? 
</div>


## Acknowledgements {.larger}

<span style="color:blue">Materials above are adapted from the following sources:</span>


- <span style="color:forestgreen">Colin Rundel's notes from Duke University</span>
- [`dplyr` and `tidyr` cheatsheets](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
- [Mine Cetinkaya-Rundel's DataFest 2016 Visualization Workshop](https://github.com/mine-cetinkaya-rundel/df2016_workshops/tree/master/viz_ggplot2_shiny)
- [Hadley Wickham's R4DS](https://r4ds.had.co.nz/data-visualisation.html#aesthetic-mappings)
- [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html)
- <span style="color:forestgreen">Chuck Lanfear's CSSS508 course at UW</span>
- Hadley Wickham - [Advanced R](http://adv-r.had.co.nz/)

