---
title: "In-class Exercises Week 5"
author: "David Post"
date: "5/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set


library(tidyr)
library(dplyr)
library(purrr)
library(magrittr)
library(ggplot2)
library(shiny)
library(ggthemes)
library(nycflights13)
```


## In-class Exercises  {.smaller .build}

1. How many flights to Los Angeles (LAX) did each of the legacy carriers (AA, UA, DL or US) have in May from JFK, and what was their average duration?

2. What was the shortest flight out of each airport in terms of distance? In terms of duration?

3. Which plane (check the tail number) flew out of each New York airport the most?

<div style="color:forestgreen">
*Optional*

Which date should you fly on if you want to have the lowest possible average departure delay? What about arrival delay? 
</div>

### 1.

We bring in our data above in the library, then we want to find the number of flights with the
required conditions.

```{r pipes and pipes}
flights %>% 
  filter( carrier == "AA" | carrier == "UA" | carrier == "DL" | carrier== "US" ) %>% 
  filter(dest == "LAX" & origin== "JFK" & month == 3 ) %>%
  summarise( flight_count = length(air_time), 
             mean_duration = mean(air_time, na.rm=T))

``` 

We see that excluding ones from the mean which did not have the flight duration listed, we have 
that there were 688 flights, averaging 334 munites, or about 5 and a half hours.

### 2.

For this one we'll want to use a `group_by` command to gather them by originating airport and
then find the minimum.
```{r short flights}
flights %>%
  group_by(origin) %>%
  summarise( min_dist = min(distance, na.rm=T),
             min_dur = min(air_time,na.rm = T))
```

Here we see that the minimum duration is 20-21 minutes, and the distance can be a short drive
from EWR or about 95 from JFK or LGA.

### 3.

```{r}
flights %>%
  filter(!is.na(tailnum)) %>%
  group_by(origin, tailnum) %>%
  summarize( dep_per_origin = length(tailnum)) %>%
  group_by(origin) %>%
  filter(dep_per_origin == max(dep_per_origin, na.rm=T))
```
